---
title: "Species Occurrence - Report"
author: "WWF Data Platform"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("source.R")
library(ggplot2)
library(gridExtra)
library(ggspatial)
library(terra)
```

```{r vars, include=FALSE}
loc <- "Pantanal" #(options: Colombia, Madagascar, Pantanal, TRIDOM)
species <- "jaguar" #(options: riverdolphin, jaguar, turtles)
pl <- "Pantanal"

##Load desired species and location information
if (loc == "Colombia"){
  loc_shp <- col_shp
} else if (loc == "Pantanal"){
  loc_shp <- pant_shp
} else if (loc == "Madagascar"){
  loc_shp <- mada_shp
} else if (loc == "TRIDOM"){
  loc_shp <- tri_shp
} else  loc_shp <- NULL

sp_shp <- st_read(paste0("clipped_data/",species,"_",loc,"_2024-11-13.shp"))
sp_table <- as.data.frame(readRDS(paste0("clipped_data/",species,"_",loc,"_2024-11-14.rds")))
baseline <- 1999

#Set baseline
sp_table <- filter(sp_table, year > baseline)
```
## Species Occurrence Report for ``r loc``

This summarizes the reported occurrences of identified species of interest for the defined area of interest. Species occurrence reports are retrieved from the Global Biodiversity Information Facility (GBIF), which aggregates and harmonizes species occurrence records from multiple sources includng museum specimens, observations from citizen science and research, as well as machine observations (e.g. camera traps). For more details GBIF and underlying data sources, please see <https://www.gbif.org/what-is-gbifm>.

This report summarizes the following species occurrence indicators for measuring Biodiversity in the ``r pl`` in ``r loc``:
- Indicator 1. Annual reported focal species occurrences of species within the area
- Indicator 2. Basis of record for all species occurrences
- Indicator 3. Number of WWF focal species reported present in the area, annually

It also provides a map of the distribution of all species occurrences records for the area.

```{r analysis, include=FALSE}
##Summarize table of number of occurrences per year
annual_occ <- sp_table %>% 
  select(gbifID,species,year,indvdlC) %>%
  distinct() %>%
  count(year,species)

occ_basis <- sp_table %>%
  select(gbifID,species,bssOfRc) %>%
  distinct() %>%
  count(bssOfRc,species) %>%
  group_by(species) %>%
  mutate(perc=n/sum(n)) %>%
  mutate(labels=scales::percent(perc))

occ_basis$bssOfRc <- gsub("_"," ",occ_basis$bssOfRc)

occ_basis <- occ_basis %>%
  mutate(bssOfRc = stringr::str_to_title(bssOfRc))

##Get positions for labels with lines
# label_df <- occ_basis %>%
#   mutate(csum = rev(cumsum(rev(n))),
#          pos = n/2 + lead(csum,1),
#          pos = if_else(is.na(pos), n/2, pos))

##Plot # of occurrences per year
plot1 <- ggplot(annual_occ, aes(x=year, y=n)) +
  geom_point(aes(colour=factor(species)),size=3, alpha=0.75, position = position_jitter(width=0.2, height=0)) +
  scale_x_continuous(name="Year",limits=c(2000,2024),breaks=c(2000:2024)) +
  scale_y_continuous(name="Occurrences",limits=c(0,plyr::round_any(max(annual_occ$n),10, f=ceiling))) +
  labs(colour="Species",
       caption = "Data source: GBIF") +
  theme_minimal() +
  theme(
    text = element_text(family="sans"),
    axis.text.x = element_text(angle=45, hjust=1),
    panel.border = element_rect(colour="darkgray",fill=NA,size=0.5),
    legend.text = element_text(face="italic"),
    axis.title.x = element_text(size="12"),
    axis.title.y = element_text(size="12"),
    plot.title = element_text(size="14", colour="#3d3a3a"),
    plot.subtitle = element_text(face="italic", colour="#3d3a3a"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill="white"),
    plot.background = element_rect(fill="white")) +
  ggtitle("OCCURRENCES PER YEAR", subtitle = paste0("Total occurrence records: ", n_distinct(sp_table$gbifID)))

##Plot # of occurrences per basis of record
plot2 <- ggplot(occ_basis, aes(x="",y=perc,fill=bssOfRc)) +
  geom_col(color="darkgray") +
  geom_label(aes(label=labels),
            position=position_stack(vjust=0.5),
            show.legend = FALSE) +
  coord_polar(theta="y") +
  scale_fill_brewer() +
  labs(fill="Basis of record",
       caption = "Data source: GBIF") +
  theme_void() +
  facet_grid(.~species, switch="both") +
  theme(
    text = element_text(family="sans"),
    legend.title = element_text(size="12"),
    legend.text = element_text(size="10"),
    plot.title = element_text(size="14", colour="#3d3a3a"),
    plot.subtitle = element_text(face="italic", colour="#3d3a3a"),
    strip.text.x = element_text(size="12",face="italic",vjust=1),
    panel.background = element_rect(fill="white",colour="white"),
    plot.background = element_rect(fill="white",colour="white"),
    plot.margin = margin(5,5,20,5)) +
  ggtitle ("OCCURRENCES PER BASIS OF RECORD", subtitle = paste0("Total occurrence records: ", n_distinct(sp_table$gbifID)))

#map by species
map <- ggplot() +
  geom_sf(data = loc_shp, fill = NA, color = "darkblue", linewidth = 0.5) +
  geom_sf(data = sp_shp,
          aes(color = species),
          size = 2,
          alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  annotation_scale(location = "bl") +
  annotation_north_arrow(location = "tr",
                         pad_x = unit(0.2, "in"),
                         pad_y = unit(0.2, "in")) +
  labs(title = paste0("DISTRIBUTION OF ",toupper(species)," OCCURRENCE RECORDS"),
       subtitle = paste("Total occurence records: ", n_distinct(sp_shp$gbifID)),
       color = "Species",
       caption = "Data source: GBIF") +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(size = 10)
  )
```
### Distribution of species occurrence in area

```{r map, echo=FALSE}
plot(map)
```

<span style="color:darkblue;">**Questions:**</span>
- Question 1
- Question 2

### Annual trends for focal species

```{r annual, echo=FALSE}
plot(plot1)
```

<span style="color:darkblue;">**Questions:**</span>
- Question 1
- Question 2

### Basis of record of occurrences

```{r basis, echo=FALSE}
plot(plot2)
```

<span style="color:darkblue;">**Questions:**</span>
- Question 1
- Question 2

### WWF focal species present in the area

The following WWF focal species have been reported in the area since 2000:

INSERT DYNAMIC TABLE

```{r wwf, echo=FALSE}

```

<span style="color:darkblue;">**Questions:**</span>
- Question 1
- Question 2

### Citations

List citations for GBIF data used